---
title: 'Bios 6301: Assignment 5'
author:
  - Erin Fey
output: pdf_document
---
Due Tuesday, 15 November, 1:00 PM

$5^{n=day}$ points taken off for each day late.

50 points total.

Submit a single knitr file (named homework5.rmd), along with a valid PDF output file. Inside the file, clearly indicate which parts of your responses go with which problems (you may use the original homework document as a template). Add your name as author to the file's metadata section. Raw R code/output or word processor files are not acceptable.

Failure to name file homework5.rmd or include author name may result in 5 points taken off.

Question 1

24 points

Import the HAART dataset (haart.csv) from the GitHub repository into R, and perform the following manipulations: (4 points each)
```{r}
library(lubridate)
haart <- read.csv("https://raw.githubusercontent.com/fonnesbeck/Bios6301/master/datasets/haart.csv")
haart[,'init.date']<- as.Date(haart[,'init.date'], format="%m/%d/%y")
haart[,'date.death']<- as.Date(haart[,'date.death'], format="%m/%d/%y")
haart[,'last.visit']<- as.Date(haart[,'last.visit'], format="%m/%d/%y")
```
Convert date columns into a usable (for analysis) format. Use the table command to display the counts of the year from init.date.
```{r}
haart[,'init.year']<-format(haart[,'init.date'],'%Y')
table(haart[,'init.year'])
```
Create an indicator variable (one which takes the values 0 or 1 only) to represent death within 1 year of the initial visit. How many observations died in year 1?
```{r}
haart[, 'death1'] <- ifelse((haart[, 'date.death'] - haart[, 'init.date'] > 365 | is.na(haart[, 'date.death'])),0,1)
sum(haart[, 'death1']==1)
```
92 patients within 1 year

Use the init.date, last.visit and death.date columns to calculate a followup time (in days), which is the difference between the first and either the last visit or a death event (whichever comes first). If these times are longer than 1 year, censor them (this means if the value is above 365, set followup to 365). Print the quantile for this new variable.
```{r}
haart[, 'follow.up'] <- ifelse(is.na(haart[, 'last.visit']), haart[, 'date.death'] - haart[, 'init.date'], haart[, 'last.visit'] - haart[, 'init.date'])
haart[, 'follow.up'][haart[, 'follow.up'] > 365] <- 365
quantile(haart[, 'follow.up'])
```
Create another indicator variable representing loss to followup; this means the observation is not known to be dead but does not have any followup visits after the first year. How many records are lost-to-followup?
```{r}
haart[,'lost'] <- ifelse(haart[,'death']==0 & haart[,'follow.up']==365,1,0)
table(haart[,'lost'])
```
710 records lost to follow up

Recall our work in class, which separated the init.reg field into a set of indicator variables, one for each unique drug. Create these fields and append them to the database as new columns. Which drug regimen are found over 100 times?
```{r}
reg_list <- strsplit(as.character(haart[,'init.reg']),',')
all_drugs <- unique(unlist(reg_list))
reg_drugs <- matrix(nrow=nrow(haart), ncol=length(all_drugs))
for(i in seq_along(all_drugs)){
    reg_drugs[,i] <- +sapply(reg_list, function(x) all_drugs[i] %in% x)
}
colnames(reg_drugs) <- all_drugs
haart <- cbind(haart, reg_drugs)
reg_drugs<- as.data.frame(reg_drugs)
sapply(reg_drugs, sum)
```
3TC, AZT, EFV, NVP, and D4T

The dataset haart2.csv contains a few additional observations for the same study. Import these and append them to your master dataset (if you were smart about how you coded the previous steps, cleaning the additional observations should be easy!). Show the first five records and the last five records of the complete (and clean) data set.
```{r}
haart <- data.frame(read.csv("https://raw.githubusercontent.com/fonnesbeck/Bios6301/master/datasets/haart.csv"))
haart2 <- data.frame(read.csv("https://raw.githubusercontent.com/fonnesbeck/Bios6301/master/datasets/haart2.csv"))
haart <- rbind(haart, haart2)
haart[,'init.date']<- as.Date(haart[,'init.date'], format="%m/%d/%y")
haart[,'date.death']<- as.Date(haart[,'date.death'], format="%m/%d/%y")
haart[,'last.visit']<- as.Date(haart[,'last.visit'], format="%m/%d/%y")
haart[, 'death1'] <- ifelse((haart[, 'date.death'] - haart[, 'init.date'] > 365 | is.na(haart[, 'date.death'])),0,1)
haart[, 'follow.up'] <- ifelse(is.na(haart[, 'last.visit']), haart[, 'date.death'] - haart[, 'init.date'], haart[, 'last.visit'] - haart[, 'init.date'])
haart[, 'follow.up'][haart[, 'follow.up'] > 365] <- 365
haart[,'lost'] <- ifelse(haart[,'death']==0 & haart[,'follow.up']==365,1,0)
reg_list <- strsplit(as.character(haart[,'init.reg']),',')
all_drugs <- unique(unlist(reg_list))
reg_drugs <- matrix(nrow=nrow(haart), ncol=length(all_drugs))
for(i in seq_along(all_drugs)){
    reg_drugs[,i] <- +sapply(reg_list, function(x) all_drugs[i] %in% x)
}
colnames(reg_drugs) <- all_drugs
haart <- cbind(haart, reg_drugs)
head(haart, n=5)
tail(haart, n=5)
```

Question 2

14 points

Use the following code to generate data for patients with repeated measures of A1C (a test for levels of blood glucose).
```{r}
genData <- function(n) {
if(exists(".Random.seed", envir = .GlobalEnv)) {
save.seed <- get(".Random.seed", envir= .GlobalEnv)
on.exit(assign(".Random.seed", save.seed, envir = .GlobalEnv))
} else {
on.exit(rm(".Random.seed", envir = .GlobalEnv))
}
set.seed(n)
subj <- ceiling(n / 10)
id <- sample(subj, n, replace=TRUE)
times <- as.integer(difftime(as.POSIXct("2005-01-01"), as.POSIXct("2000-01-01"), units='secs'))
dt <- as.POSIXct(sample(times, n), origin='2000-01-01')
mu <- runif(subj, 4, 10)
a1c <- unsplit(mapply(rnorm, tabulate(id), mu, SIMPLIFY=FALSE), id)
data.frame(id, dt, a1c)
}
x <- genData(500)
```
Perform the following manipulations: (2 points each)

Order the data set by id and dt.
```{r}
x <- x[order(x$id,x$dt),]
```
For each id, determine if there is more than a one year gap in between observations. Add a new row at the one year mark, with the a1c value set to missing. A two year gap would require two new rows, and so forth.
```{r}
gap.check <- function(identity,date){
  insert <- vector()
  new.row <- vector()
  for (i in unique(identity)){
    rows <- which(identity==i)[1:length(which(identity==i))-1]
    for (j in rows){
      new.row <- c(new.row, j)
      if(unclass(difftime(date[j+1], date[j], "days"))[1] > 366){
        insert <- c(insert,j+1)
      }
    }
  }
  return(insert)
}

add.row <- function(df,insertion){
    df <- rbind(df[1:(insertion-1),],data.frame(id=df$id[insertion-1],
                                              dt=df$dt[insertion-1]+years(1),a1c=NA),
                df[insertion:nrow(df),])
  return(df)
}

p <- x
insert <- gap.check(p$id,p$dt)
lines <- insert+seq(from=0,by=1,length.out=length(insert))
for (i in 1:length(lines)){
  p <- add.row(p,lines[i])
}

(insert <- gap.check(p$id,p$dt))
x <- p
```
Create a new column visit. For each id, add the visit number. This should be 1 to n where n is the number of observations for an individual. This should include the observations created with missing a1c values.
```{r}
for (i in 1:length(unique(x$id))){
  visit <- seq(1:table(x$id)[[i]])
  x$visit[x$id==i] <- visit
}
```
For each id, replace missing values with the mean a1c value for that individual.
```{r}
for (i in 1:length(unique(x$id))){
  rows <- which(x$id==i)
  meana1c <- mean(x$a1c[rows[1]:tail(rows,n=1)],na.rm = TRUE)
  for (j in rows){
    if(is.na(x$a1c[j])){
      x$a1c[j] <- meana1c
    }
  }
}
```
Print mean a1c for each id.
```{r}
for (i in 1:length(unique(x$id))){
  rows <- which(x$id==i)
  meana1c <- mean(x$a1c[rows[1]:tail(rows,n=1)])
  print(c(as.integer(i),meana1c))
} 
```
Print total number of visits for each id.
```{r}
table(x$id)
```
Print the observations for id = 15.
```{r}
x[which(x$id==15),]
```

Question 3

10 points

Import the addr.txt file from the GitHub repository. This file contains a listing of names and addresses (thanks google). Parse each line to create a data.frame with the following columns: lastname, firstname, streetno, streetname, city, state, zip. Keep middle initials or abbreviated names in the firstname column. Print out the entire data.frame.
```{r}
addr <- read.delim("https://raw.githubusercontent.com/fonnesbeck/Bios6301/master/datasets/addr.txt", stringsAsFactors = FALSE, head=FALSE)
temp<-unlist(strsplit(addr[,1],"  "))
trim <- function (x) gsub("^\\s+|\\s+$", "", x)
temp<-trim(temp)
temp<-temp[temp!=""]
mt<-matrix(temp,ncol=6,byrow=T)
rexp <- "^(\\w+)\\s?(.*)$"
y <- data.frame(streetno=sub(rexp,"\\1",mt[,3]), streetname=sub(rexp,"\\2",mt[,3]))
mt<-cbind(y,mt)
df<-as.data.frame(mt[,-5])
colnames(df)<-c("streetno", "streetname", "lastname", "firstname", "city", "state", "zip")
df<-df[,c(3,4,1,2,5,6,7)]
print(df)
```

Question 4

2 points

The first argument to most functions that fit linear models are formulas. The following example defines the response variable death and allows the model to incorporate all other variables as terms. . is used to mean all columns not otherwise in the formula.
```{r}
url <- "https://github.com/fonnesbeck/Bios6301/raw/master/datasets/haart.csv"
haart_df <- read.csv(url)[,c('death','weight','hemoglobin','cd4baseline')]
coef(summary(glm(death ~ ., data=haart_df, family=binomial(logit))))
```
Now imagine running the above several times, but with a different response and data set each time. Here's a function:
```{r}
  myfun <- function(dat, response) {
    form <- as.formula(response ~ .)
    coef(summary(glm(form, data=dat, family=binomial(logit))))
  }
```
  Unfortunately, it doesn't work. tryCatch is "catching" the error so that this file can be knit to PDF.
```{r}
  tryCatch(myfun(haart_df, death), error = function(e) e)
```
  What do you think is going on? Consider using debug to trace the problem.
When you use death for the response, it looks for the stored object death which is not a variable

5 bonus points
  
Create a working function.
```{r}
myfun_1 <- function(dat, response) {
  form <- as.formula(paste(response, "~."))
  coef(summary(glm(form, data=dat, family=binomial(logit))))
}
myfun_1(haart_df, 'death')
```